Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-02-05 article.tex
modification : typo modification : readability in some proof

\begin{proof} 
We prove the lemma by induction on the size of the motif. Suppose that we have the motif $(12)$, thus the stripe is $\ustep$. If there is a match in stripe, then there exists two (contiguous) increasing elements in the text, which is a match in permutation. Same goes for the permutation $(21)$. Now, suppose that the lemma is true for every motif of size $n$. Let $\pmotif$ be a motif of size $n+1$ avoiding $(213,231)$, and $\ptext$ a permutation avoiding $(213,231)$. If there is a match of $\stripew{\pmotif}$ in $\stripew{\ptext}$ by subsequence $s$, then there is a match of $\stripew{\pmotif}[2:]$ in $\stripew{\ptext}[2:]$ by subsequence $s[2:]$, and by induction there is a match of $\pmotif[2:]$ in $\ptext[2:]$ by subsequence $s[2:]$. Moreover suppose that the first element the stripe of the motif is ascend (descent), and by lemma \ref{lemma:first element is 1 or n} the minimal (maximal) element, and remark that a subsequence of a permutation avoiding (231,213) is also avoiding (231,213), thus the first element of the subsequence $s$ is either the minimal (maximal) element which complete our proof.
\end{proof}

@@

\textbf{Proposition  \num \thenum} Let $\pmotif$ and $\ptext$
be two permutations avoiding (231,213), 
we can decide in linear time if $\pmotif$ 
appear in $\ptext$.\\

@@

\begin{proof}
Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$. The optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element $j'$ of $\ptext[j:]$ such that : (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1). Indeed when we will match $\factor{i+1}\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$, every elements of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$. Remark that the maximal (minimal) element of $\factor{i}\factor{i-1}\cdots\factor{1}$ is $\first{i-1}$.\\

@@

		MIN_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } & 
		\text{if $\stripe{\pmotif}{\first{1}}=\ustep$}\\
		\text{ $LIS(j,j',\ptext[j']) \leq \factor{1}$}\}&\\

@@

		MAX_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } &
		\text{if $\stripe{\pmotif}{\first{1}}=\dstep$} \\
		\text{ $LDS(j,j',\ptext[j']) \leq \factor{1}$}\}&\\

@@

$I(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $LIS(j,j'-1,\first{i-1}) \leq |\factor{i}| \}$ \\

$D(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $LDS(j,j'-1,\first{i-1}) \leq |\factor{i}| \}$ \\

@@

In the base case, 
we are looking for a match for the first factor.
Each factor is either ascent or descent.
If the factor is an ascent (descent)
we have to find an increasing (decreasing) subsequence
in the text of same size or longest that 
the size of the factor,
and to find the optimal solution 
we must assure that the last element
of the sequence is minimal (maximal).\\
For the induction, it is the same idea
except that we must assure that 
every element of the subsequence is superior 
to the maximum element of the rest of the match
ie inferior to the minimal (maximal) element of the rest of the match
which is given by $LM$ of
the previous factor.\\

@@

			if $n_\pmotif \notin  \x : $\\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,j)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[j]<\max$
						}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,j)= true $\\

			if $n_\pmotif \in  \x : $\\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,n_\ptext)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[n_\ptext]<\max$
						}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,n_\ptext)= true $\\

			if $n_\pmotif \notin  \x$  
			and $n_\pmotif \in  \y$ 
			and $\pmotif[n_\pmotif]$ is the maximal element: \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,j)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[j]<\max$
						}\\
						&\text{$\ptext[j]$ is the maximal element}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,j)= true $\\

			if $n_\pmotif \in  \x$  
			and $n_\pmotif \in  \y$ 
			and $\pmotif[n_\pmotif]$ is the maximal element: \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,n_\ptext)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[n_\ptext]<\max$
						}\\
						&\text{$\ptext[n_\ptext]$ is the maximal element}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,n_\ptext)= true $\\

@@

						\mathbb{P}_{\bmotif,\ptext,t(j),\max}(i+1,j+1) 
							& \text{if $\stripe{\pmotif}{i}=\ustep$}\\
							& \text{and $i \in \x$ and $i \notin \y$}\\
							& \text{and $\min<\ptext[j]<\max$} \\
						&\\

						\mathbb{P}_{\bmotif,\ptext,\min,t(j)}(i+1,j+1) 
							& \text{if $\stripe{\pmotif}{i}=\dstep$}\\
							& \text{and $i \in \x$ and $i \notin \y$ }\\
							& \text{and $\min<\ptext[j]<\max$} \\

@@

						\mathbb{P}_{\bmotif,\ptext,\ptext[j]),\max}(i+1,j+1) 
							&\text{if $\stripe{\pmotif}{i}=\ustep$}\\
							&\text{and $i \in \x$ and $i \in \y$}\\ 
							&\text{and $\ptext[j]+1=\ptext[j+1]$ }\\
							& \text{and $\min<\ptext[j]<\max$} \\
						&\\

@@

			If $i \in \x$, $i \in \y$ and i is a descent,
			we match $\pmotif[i]$ to $\ptext[j]$, 
			then we match $\pmotif[i+1:]$ with $\ptext[j+1:]$
			with the condition that
			$\pmotif[i+1]$ is matched to $\ptext[j+1]$,
			and that $\ptext[j+1]=\pmotif[i]-1$.\\

			If $i \in \x$, $i \in \y$ and i is an ascent,
			we match $\pmotif[i]$ to $\ptext[j]$, 
			then we match $\pmotif[i+1:]$ with $\ptext[j+1:]$
			with the condition that
			$\pmotif[i+1]$ is matched to $\ptext[j+1]$,
			and that $\ptext[j+1]=\pmotif[i]+1$.\\

@@

	\begin{proof}
	Let's $s$ be a longest subsequence avoiding (213,231) with last element at index $f$ in $\pi$, 
	suppose that $P(\pi)$ is not a longest increasing subsequence with last element at index $f$. Let's $s_m$ be a longest increasing subsequence with last element $f$.
	Thus $|s_m|>P(\pi)$, clearly the sequence $s_m \cup M(\pi)$
	is avoiding (213,231) and is longer than $s$ witch is not possible. 
	The case for $M(\pi)$ follows the same idea.
	\end{proof}

@@

	\begin{proposition}
	Given two permutation $\pi_1$ and $\pi_2$
	The longest common subsequence
	avoiding (231,213) can be solve in $O(|\pi_1|^3|\pi_2|^3)$.	
	\end{proposition}

@@

	$S_{\pi_1,\pi_2}(\min_1,\max_1,\min_2,\min_2,i_1,i_2)$
	= max$\{$ $|m|$ $|$ 
	s is a pattern occurring 
	in  $\pi_1[i_1:]$
	by the subsequence $s_1$ and $\min(s_1)=\min_1$ and $\max(s_1)=\max_1$
	and s is occurring
	in  $\pi_2[i_2:]$
	by the subsequence $s_2$ and $\min(s_2)=\min_2$ and $\max(s_2)=\max_2$	
	$\}$\\

@@

	\textbf{BASE :} \\
	$S_{\pi_1,\pi_2,\min_1,\max_1,\min_2,\max_2}(|\pi_1|,|\pi_2|) = \left\{ 
			\begin{array}{l l}
				1 & \text{if $\min_1<\pi_1[j]<\max_1$  
				}\\
				& \text{ and $\min_2<\pi_2[j]<\max_2$}\\
				0 & otherwise\\
			\end{array} \right. $\\

@@

	with 
	$S'_{\pi_1,\pi_2,\min_1,\max_1,\min_2,\max_2}(i_1,i_2)=\\ \left\{ 
				\begin{array}{l l}
					1+S_{\pi_1,\pi_2,\pi_1[i_1],\max_1,\pi_2[i_2]}(\max_2,i_1+1,i_2+1)  & \text{$\pi_1[i_1]<\min_1$ } \\
					& \text{and $\pi_2[i_2]<\min_2$} \\
					&\\
					1+S_{\pi_1,\pi_2,\min_1,\pi_1[i_1],\min_2,\pi_2[i_2]}(i_1+1,i_2+1)  & \text{$\pi_1[i_1]>\max_1$ } \\	
					&\text{and $\pi_2[i_2]>\max_2$}\\	
					&\\	
					0 & \text{otherwise}\\							
				\end{array} \right.$\\

>>>>

\begin{proof} 
We prove the lemma by induction on the size of the motif.The base case is a motif of size 2. Suppose that we have the motif $(12)$, thus the stripe of the motif is $\ustep$. If there is a match in stripe, then there exists two (contiguous) increasing elements in the text, which is a match in permutation. Same goes for the permutation $(21)$. Now, suppose that the lemma is true for every motif in $\Av{n}{(231,213)}$. Let $\pmotif$ be a motif in $\Av{n+1}{(231,213)}$, and $\ptext$ a permutation avoiding $(213,231)$. If there is a match of $\stripew{\pmotif}$ in $\stripew{\ptext}$ by subsequence $s$, then there is a match of $\stripew{\pmotif}[2:]$ in $\stripew{\ptext}[2:]$ by subsequence $s[2:]$, and by induction there is a match of $\pmotif[2:]$ in $\ptext[2:]$ by subsequence $s[2:]$. Moreover if the first element the stripe of the motif is ascend (descent), and by lemma \ref{lemma:first element is 1 or n} the minimal (maximal) element, and remark that a subsequence of a permutation avoiding (231,213) is also avoiding (231,213), thus the first element of the subsequence $s$ is either the minimal (maximal) element which complete our proof.
\end{proof}

@@

\begin{proposition}
Let $\pmotif$ and $\ptext$
be two permutations avoiding (231,213), 
we can decide in linear time if $\pmotif$ 
appear in $\ptext$.\\	
\end{proposition}

@@

\begin{proof}
Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, the optimal match is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$, the optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element at index $j'$ of $\ptext[j:]$ such that : (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1). Indeed when we will match $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$, every element of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$.

\begin{remark}
Given an ascent (descent) factor $\factor{i}$, the maximal (minimal) element of $\factor{i}\factor{i-1}\cdots\factor{1}$ is $\first{i-1}$.			  
\end{remark}

\begin{remark}
An ascent (descent) factor 	$\factor{i}$ is a match in $\ptext[j:j']$  if and only if there exists an increasing (decreasing) subsequence in $\ptext[j:j']$ of size superior or equal to the size of $\factor{i}$.		  
\end{remark}

@@

		MIN_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } & 
		\text{if $\stripe{\pmotif}{\first{1}}=\ustep$}\\
		\text{ $|\factor{1}| \leq LIS(j,j',\ptext[j'])  $}\}&\\

@@

		MAX_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } &
		\text{if $\stripe{\pmotif}{\first{1}}=\dstep$} \\
		\text{ $|\factor{1}| \leq LDS(j,j',\ptext[j']) $}\}&\\

@@

$I(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $|\factor{i}| \leq LIS(j,j'-1,\first{i-1})\}$ \\

$D(i,j)=\{j'|\text{$j<j'$ and $LM(i-1,j')\neq 0$}$ \\
\indent \indent \indent and  $|\factor{i}| \leq LDS(j,j'-1,\first{i-1}) \}$ \\

@@

In the base case, 
we are looking for a match for the first factor.
Each factor is either ascent or descent.
If the factor is an ascent (descent)
we have to find an increasing (decreasing) subsequence
in the text of same size or longest that 
the size of the factor,
and to find the optimal solution 
we must assure that the last element
of the sequence is minimal (maximal).\\
For the induction, it is the same idea
except that we must assure that 
every element of the subsequence is superior (inferior)
to the maximum (minimal) element of the rest of the match
ie inferior (superior) to the minimal (maximal) element of the rest of the match
which is given by $LM$ of
the previous factor.\\

@@

			if $n_\pmotif \notin  \x $
			and $n_\pmotif \notin  \y$ : \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,j)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[j]<\max$
						}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,j)= true $\\

			if $n_\pmotif \in  \x $
			and $n_\pmotif \notin  \y$ : \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,n_\ptext)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[n_\ptext]<\max$
						}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,n_\ptext)= true $\\

			if $n_\pmotif \notin  \x$  
			and $n_\pmotif \in  \y$ 
			and $\pmotif[n_\pmotif]$ is the maximal element: \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,j)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[j]<\max$
						and }\\
						&\text{$\ptext[j]$ is the maximal element}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,j)= true $\\

			if $n_\pmotif \in  \x$  
			and $n_\pmotif \in  \y$ 
			and $\pmotif[n_\pmotif]$ is the maximal element: \\		
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif,n_\ptext)= \left\{ 
					\begin{array}{l l}
						true & \text{if $\min<\ptext[n_\ptext]<\max$
						and }\\
						&\text{$\ptext[n_\ptext]$ is the maximal element}\\
						false & otherwise\\
					\end{array} \right. $\\	
			$\mathbb{P}_{\bmotif,\ptext,\min,\max}(n_\pmotif+1,n_\ptext)= true $\\

@@

						\mathbb{P}_{\bmotif,\ptext,\ptext[j],\max}(i+1,j+1) 
							& \text{if $\stripe{\pmotif}{i}=\ustep$}\\
							& \text{and $i \in \x$ and $i \notin \y$}\\
							& \text{and $\min<\ptext[j]<\max$} \\
						&\\

						\mathbb{P}_{\bmotif,\ptext,\min,\ptext[j]}(i+1,j+1) 
							& \text{if $\stripe{\pmotif}{i}=\dstep$}\\
							& \text{and $i \in \x$ and $i \notin \y$ }\\
							& \text{and $\min<\ptext[j]<\max$} \\

@@

						\mathbb{P}_{\bmotif,\ptext,\ptext[j],\max}(i+1,j+1) 
							&\text{if $\stripe{\pmotif}{i}=\ustep$}\\
							&\text{and $i \in \x$ and $i \in \y$}\\ 
							&\text{and $\ptext[j]+1=\ptext[j+1]$ }\\
							& \text{and $\min<\ptext[j]<\max$} \\
						&\\

@@

			If $i \in \x$, $i \in \y$ and i is a descent,
			we match $\pmotif[i]$ to $\ptext[j]$, 
			then we match $\pmotif[i+1:]$ with $\ptext[j+1:]$
			with the condition that
			$\pmotif[i+1]$ is matched to $\ptext[j+1]$,
			and that $\ptext[j+1]=\ptext[j]-1$.\\

			If $i \in \x$, $i \in \y$ and i is an ascent,
			we match $\pmotif[i]$ to $\ptext[j]$, 
			then we match $\pmotif[i+1:]$ with $\ptext[j+1:]$
			with the condition that
			$\pmotif[i+1]$ is matched to $\ptext[j+1]$,
			and that $\ptext[j+1]=\ptext[j]+1$.\\

@@

	\begin{proof}
	Let's $s$ be a longest subsequence avoiding (213,231) with last element at index $f$ in $\pi$, 
	suppose that $P(\pi)$ is not a longest increasing subsequence with last element at index $f$. Let's $s_m$ be a longest increasing subsequence with last element $f$.
	Thus $|s_m|>|P(\pi)|$, clearly the sequence $s_m \cup M(\pi)$
	is avoiding (213,231) and is longer than $s$ witch is not possible. 
	The same idea goes for the demonstration that $M(\pi)$ is the longest decreasing subsequence.
	\end{proof}

@@

	\begin{proposition}
	Given two permutation $\pi_1$ and $\pi_2$,
	the longest common subsequence
	avoiding (231,213) can be compute in $O(|\pi_1|^3|\pi_2|^3)$ time and space.	
	\end{proposition}

@@

	$S_{\pi_1,\pi_2}(\min_1,\max_1,\min_2,\min_2,i_1,i_2)$
	= max$\{$ $|s|$ $|$ 
	s is a pattern occurring 
	in  $\pi_1[i_1:]$
	by the subsequence $s_1$ and $\min(s_1)=\min_1$ and $\max(s_1)=\max_1$
	and s is occurring
	in  $\pi_2[i_2:]$
	by the subsequence $s_2$ and $\min(s_2)=\min_2$ and $\max(s_2)=\max_2$	
	$\}$\\

@@

	\textbf{BASE :} \\
	$S_{\pi_1,\pi_2,\min_1,\max_1,\min_2,\max_2}(|\pi_1|,|\pi_2|) = \left\{ 
			\begin{array}{l l}
				1 & \text{if $\min_1=\pi_1[j]=\max_1$  
				}\\
				& \text{ and $\min_2=\pi_2[j]=\max_2$}\\
				0 & otherwise\\
			\end{array} \right. $\\

@@

	with 
	$S'_{\pi_1,\pi_2,\min_1,\max_1,\min_2,\max_2}(i_1,i_2)=\\ \left\{ 
				\begin{array}{l l}
					1+S_{\pi_1,\pi_2,\pi_1[i_1],\max_1,\pi_2[i_2]}(i_1+1,i_2+1)  & \text{$\pi_1[i_1]<\min_1$ } \\
					& \text{and $\pi_2[i_2]<\min_2$} \\
					&\\
					1+S_{\pi_1,\pi_2,\min_1,\pi_1[i_1],\min_2,\pi_2[i_2]}(i_1+1,i_2+1)  & \text{$\pi_1[i_1]>\max_1$ } \\	
					&\text{and $\pi_2[i_2]>\max_2$}\\	
					&\\	
					0 & \text{otherwise}\\							
				\end{array} \right.$\\
